---
title: "Public health and economic casulaties due to meteoroligical events in the United States between 1950 and 2011"
author: "Akim van Eersel"
date: "2020-09-27"
output: 
  html_document: 
    keep_md: yes
    fig_width: 9
    fig_height: 8
editor_options: 
  chunk_output_type: console
---
# Synopsis

In this report we aim to describe most impactful population health consequences and economic damages casualties made by storms and other weather events in the United States between 1950 and 2011. From the U.S. National Oceanic and Atmospheric Administration's (NOAA) database we can access the storm data set in order to answer questions about severe weather events. Impactful casualty is defined here as the biggest amount of people touched or economic damages by one weather event. From this setup, for population health consequences it was one tornado with 1742 casualties which was the most harmful event. For economic damages it was one flood with 115 032 500 000$ of damage in total which was the most impactful event. However, two other definitions are explored in this report and results are greatly distinct from previous. One is the proportion of types of weather events impacting population health or economy. The other is the total amount of casualties through all types of events occurred.

# System info
```{r systeminfo}
## Overview of session
sessionInfo()

## Rstudio version
rstudioapi::versionInfo()$version

## Files in working directory
list.files(getwd())
```


# Get, Load and Process the Raw Data

## Download data set

Download and store the data file :
```{r getdata}
## Create a folder to store data set
if (!file.exists("data")){
      dir.create("data")
}

## Download the data set from url source
if (!file.exists("data/Storm Data.bz2")){
      download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                    destfile = "data/Storm Data.bz2")
}

## Check downloading date and file
dateDownloaded <- date()
dateDownloaded
list.files("data")
```

## Read the data set

Load the data file, "Storm Data.bz2", previously downloaded :
```{r readcsv,cache=TRUE}
storm <- read.csv("data/Storm Data.bz2")
```

Overview of the data set :
```{r dataoverview, cache = TRUE}
# Dimensions
dim(storm)

# Variables names
colnames(storm)

# First rows
head(storm, n = 3)
```

# Insights on variables of interest

## Problematic

The present analysis address the following questions :  
  
      1. Across the United States, which types of events are most harmful with respect to population health ?  
      2. Across the United States, which types of events have the greatest economic consequences ?  
      
* To answer these questions, the common variable of interest is "EVTYPE" for types of events on both questions.  
* Question n°1 deals with "FATALITIES" and "INJURIES" variables to know population health harmful consequences related to each events.  
* Answering the secondary question needs "PROPDMG" and "CROPDMG" variables along with theirs related decimal unit prefix stored in "PROPDMGEXP" and "CROPDMGEXP" variables. These two variables point to the economic amount casualties.

## Analysis

### Population health consequences

#### Data processing

Subset to get only the working variables :
```{r health.subset, message=FALSE}
library(dplyr)

health <- storm[, c("EVTYPE", "FATALITIES", "INJURIES")]
```

Add fatalities and injuries variables count to get total amount of population health consequences :
```{r health.casulaties, message=FALSE}
health <- health %>% mutate(Casualties = FATALITIES + INJURIES)
```

Overview of the casualties variable :
```{r health.casualties.overview, message=FALSE}
summary(health$Casualties)

# Proportion of harmless events without types distinction :
mean(health$Casualties == 0)

# Summary of harmful events without types distinction :
summary(health$Casualties[health$Casualties != 0])
```

Now, the data is grouped by types of events and calculations made to get questioning results :
```{r health.casualties.harmful}
# Group by types of events
byEvtype <- health %>% group_by(EVTYPE)

# Aggregate by types of events with their related maximum number of casualties and proportion of events with at least one casualty
summaryzero <- summarize(byEvtype, maxCasu = max(Casualties),
                         totalCasu = sum(Casualties),
                         propHarmEvent = sum(Casualties != 0)/n(),
                         numEvent = n())
```

Show top 10 of impactful events with the above processed data :
```{r health.results}
# Show the above result by descending order :
## 1. maximum number of casualties made by one event for each types of events
ansmax <- arrange(summaryzero[summaryzero$maxCasu !=0,],
                  desc(maxCasu), desc(totalCasu))
ansmax

## 2. proportion of types of events with at least one casualty
ansprop <- arrange(summaryzero[summaryzero$propHarmEvent !=0,],
                  desc(propHarmEvent), desc(maxCasu))
ansprop

## 3. total amount casualties for each types of events
anstot <- arrange(summaryzero[summaryzero$totalCasu !=0,],
                   desc(totalCasu), desc(maxCasu))
anstot
```

Since most of 10th impactful types of events are present both in summary tables of maximum and total number of casualties, let's compare the two variables with the intersecting data.  
For better comparison, a plot is made :
```{r health.tidyPlot}
library(tidyr)
library(ggplot2)
library(RColorBrewer)

# Make tidy data set to make easier plotting :
interdata <- ansmax[1:10, c("EVTYPE", "maxCasu", "totalCasu")][ansmax$EVTYPE[1:10] %in% anstot$EVTYPE[1:10],]
colnames(interdata) <- c("TypesOfEvents", "Maximum number of casualties for one event", "Total number of casualties through all observed types of events")
tidyInterdata <- gather(interdata, key = countingMethod, value = count, -TypesOfEvents)

# Plot :
pal <- colorRampPalette(brewer.pal(3,"Set2"))

g <- ggplot(tidyInterdata, aes(TypesOfEvents, count))
g + geom_col(fill = c(pal(5),pal(5))) + facet_wrap(.~countingMethod, scales = "free") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Crossing biggest impactful events types between maximum and total amount of casualties") + 
  labs(y = "Number of casualties")
```


#### Results

The above results show different types of answer:  
 
      1. One is dealing with the maximum number of casualties, "maxCasu" variable. In that sense, one **tornado** event is the most harmful with the biggest event doing 1742 casualties.  
      
      2. The second is dealing with the propotion of types of events with at least one casualty, "popHarmEvent" variable. This result is skewed since most of the bigger proportion variable values are related to few events types occurences. By consequence, "popHarmEvent" top 10 have just one of each type of event observed. In that sense, **heat wave** is the most harmful event with 70 casualties in only one apparition in the data base. Also, the data labels could be miss leading since **heat wave drought** is another type of event and can be found in the same "popHarmEvent" top 10 results, on position 7 with 19 casualties. If needed, clustering labels could avoid to many distinction between similar labels. 
      
      3. Last one is dealing with the total amount of casualties through all occurred events in the database, "totalCasu" variable. In that sense, **tornado** is again the most harmful with a total of 96 979 casualties through 60 652 occurrences.  

Since the analysis is focused on the addition of injuries and fatalities, a deeper analysis could find more precision by doing the same process with distinction between casualties. Other calculations could be made to define harmful casualties, like the total amount of casualties by events types or the related proportion.

### Economic consequences

This process analysis is mostly identical to the previous one.

#### Data processing

Subset to get only the working variables :
```{r eco.subset}
econ <- storm[, c("EVTYPE", "PROPDMG", "CROPDMG", "PROPDMGEXP", "CROPDMGEXP")]
```

With related unit prefix variables get the explicit number of economic amount casualties :
```{r eco.prefix, message=FALSE, warning=FALSE}
# Types of prefix
table(econ$PROPDMGEXP)
table(econ$CROPDMGEXP)

# Replace prefix nomination as a related decimal unit multiplication variable
multipre <- function(prefix){
      if (prefix == "" | prefix == 0){
            1
      } else if (prefix == "K" | prefix == "k") {
            10^3
      } else if (prefix == "M" | prefix == "m"){
            10^6
      } else if (prefix == "B" | prefix == "b"){
            10^9
      } else if (!is.na(as.numeric(prefix))){
            as.numeric(prefix)
      } else {
            0
      }
}
econ <- econ %>% mutate(PropMulti = sapply(PROPDMGEXP, multipre), CropMulti = sapply(CROPDMGEXP, multipre))
```

The above code is subjective to interpretation since some prefix in the data set are outside of the range specified by the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), which is quoted as :  
> Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions.   

Add fatalities and injuries variables count to get total amount of population health consequences :
```{r eco.casualties, message=FALSE}
econ <- econ %>% mutate(Casualties = PROPDMG*PropMulti + CROPDMG*CropMulti)
```

Overview of the casualties variable :
```{r eco.casualties.overview, message=FALSE}
summary(econ$Casualties)

# Proportion of harmless events :
mean(econ$Casualties == 0)

# Without harmless events :
summary(econ$Casualties[econ$Casualties != 0])
```

Now, the data is grouped by types of events and 
```{r econ.casualties.harmful}
# Group by types of events
byEvtype <- econ %>% group_by(EVTYPE)

# Aggregate by types of events with their related maximum number of casualties and proportion of events with at least one casualty
summaryzero <- summarize(byEvtype, maxCasu = max(Casualties),
                         totalCasu = sum(Casualties),
                         propHarmEvent = sum(Casualties != 0)/n(),
                         numEvent = n())
```

Show top 10 of impactful events with the above processed data :
```{r econ.results}
# Show the above result by descending order :
## 1. maximum number of casualties made by one event for each types of events
ansmax <- arrange(summaryzero[summaryzero$maxCasu !=0,],
                  desc(maxCasu), desc(totalCasu))
ansmax

## 2. proportion of types of events with at least one casualty
ansprop <- arrange(summaryzero[summaryzero$propHarmEvent !=0,],
                  desc(propHarmEvent), desc(maxCasu))
ansprop

## 3. total amount casualties for each types of events
anstot <- arrange(summaryzero[summaryzero$totalCasu !=0,],
                   desc(totalCasu), desc(maxCasu))
anstot
```

Since most of 10th impactful types of events are present both in summary tables of maximum and total number of casualties, let's compare the two variables with the intersecting data.  
For better comparison, a plot is made :
```{r econ.tidyPlot}
library(tidyr)
library(ggplot2)
library(RColorBrewer)

# Make tidy data set to make easier plotting :
interdata <- ansmax[1:10, c("EVTYPE", "maxCasu", "totalCasu")][ansmax$EVTYPE[1:10] %in% anstot$EVTYPE[1:10],]
colnames(interdata) <- c("TypesOfEvents", "Maximum number of casualties for one event", "Total number of casualties through all observed types of events")
tidyInterdata <- gather(interdata, key = countingMethod, value = count, -TypesOfEvents)

# Plot :
pal <- colorRampPalette(brewer.pal(3,"Set2"))

g <- ggplot(tidyInterdata, aes(TypesOfEvents, count))
g + geom_col(fill = c(pal(7),pal(7))) + facet_wrap(.~countingMethod, scales = "free") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
  labs(title = "Crossing biggest impactful events types between maximum and total amount of casualties") + 
  labs(y = "Number of casualties")
```

#### Results

Same as the previous section, the above results show two different types of answer:  
 
      1. One is dealing with the maximum number of dollars of damage on property and crops, "maxCasu" variable. In that sense, **flood** events are the most impactful with the biggest event doing 115 032 500 000$ of damage.  
      
      2. The other is dealing with the propotion of types of events with at least one casualty, "popHarmEvent" variable. This result is skewed since most of the bigger proportion variable values are related to few events types occurences. By consequence, "popHarmEvent" top 10 have few type of event observed. In that sense, the aggreagte **tornadoes, tstm wind, hail** is the most harmful event(s) with 1 602 500 000$ of damage.
      
            3. Last one is dealing with the total amount of economic damages through all occurred events in the database, "totalCasu" variable. In that sense, **flood** is again the most harmful with a total of 150 319 678 257$ of damage through 15 326 occurrences. It is interesting to note that 76% of this amount is due only to a single event, the one mentioned in point 1 above.

Since the analysis is focused on the addition of property and crops damages, a deeper analysis could find more precision by doing the same process with distinction between the two variables. Other calculations could be made to define harmful casualties, like the total amount of casualties by events types or the related proportion.