---
title: "Public health and economic casulaties due to severe meteoroligical events in the United States between 1950 and 2011"
author: "Akim van Eersel"
date: "2020-09-26"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---
# Synopsis

In this report we aim to describe most impactful population health consequences and economic damages casualties made by storms and other sever weather events in the United States between 1950 and 2011. From the U.S. National Oceanic and Atmospheric Administration's (NOAA) database we can access the storm data set in order to answer questions about severe weather events. Impactful casualty is defined here as the biggest amount of people touched or economic damages by a weather event. From these setup, for population health consequences it was one tornado with 1742 casualties which was the most harmful event. For economic damages it was one flood with 115 032 500 000$ of damage which was the most impactful event.

# System info
```{r systeminfo}
## Overview of session
sessionInfo()

## Rstudio version
rstudioapi::versionInfo()$version

## Files in working directory
list.files(getwd())
```


# Get, Load and Process the Raw Data

## Download data set

Download and store the data file :
```{r getdata}
## Create a folder to store data set
if (!file.exists("data")){
      dir.create("data")
}

## Download the data set from url source
if (!file.exists("data/Storm Data.bz2")){
      download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                    destfile = "data/Storm Data.bz2")
}

## Check downloading date and file
dateDownloaded <- date()
dateDownloaded
list.files("data")
```

## Read the data set

Load the data file, "Storm Data.bz2", previously downloaded :
```{r readcsv,cache=TRUE}
storm <- read.csv("data/Storm Data.bz2")
```

Overview of the data set :
```{r dataoverview, cache = TRUE}
# Dimensions
dim(storm)

# Variables names
colnames(storm)

# First rows
head(storm, n = 3)
```

# Insights on variables of interest

## Problematic

The present analysis address the following questions :  
  
      1. Across the United States, which types of events are most harmful with respect to population health ?  
      2. Across the United States, which types of events have the greatest economic consequences ?  
      
* To answer these questions, the common variable of interest is "EVTYPE" for types of events on both questions.  
* Question n°1 deals with "FATALITIES" and "INJURIES" variables to know population health harmful consequences related to each events.  
* Answering the secondary question needs "PROPDMG" and "CROPDMG" variables along with theirs related decimal unit prefix stored in "PROPDMGEXP" and "CROPDMGEXP" variables. These two variables point to the economic amount casualties.

## Results

### Population health consequences

#### Data processing

Subset to get only the working variables :
```{r health.subset, message=FALSE}
library(dplyr)

health <- storm[, c("EVTYPE", "FATALITIES", "INJURIES")]
```

Add fatalities and injuries variables count to get total amount of population health consequences :
```{r health.casulaties, message=FALSE}
health <- health %>% mutate(Casualties = FATALITIES + INJURIES)
```

Overview of the casualties variable :
```{r health.casualties.overview, message=FALSE}
summary(health$Casualties)

# Proportion of harmless events :
mean(health$Casualties == 0)

# Without harmless events :
summary(health$Casualties[health$Casualties != 0])
```

Now, the data is grouped by types of events and 
```{r health.casualties.harmful}
# Group by types of events
byEvtype <- health %>% group_by(EVTYPE)

# Aggregate by types of events with their related maximum number of casualties and proportion of events with at least one casualty
summaryzero <- summarize(byEvtype, harmful = max(Casualties),
                         propHarmEvent = sum(Casualties != 0)/n(),
                         numEvent = n())

# Show the above result by descending order by :
## 1. maximum number of casualties
arrange(summaryzero[summaryzero$harmful !=0,], desc(harmful), desc(propHarmEvent))

## 2.proportion of types of events with at least one casualty
arrange(summaryzero[summaryzero$harmful !=0,],
        desc(propHarmEvent), desc(harmful))
```

#### Conclusion

The above results show two different types of answer:  
 
      1. One is dealing with the maximum number of casualties, "harmful" variable. In that sense, **tornado** events are the most harmful with the biggest event doing 1742 casualties.  
      
      2. The other is dealing with the propotion of types of events with at least one casualty, "popHarmEvent" variable. This result is skewed since most of the bigger proportion variable values are related to few events types occurences. By consequence, "popHarmEvent" top 10 have just one type of event observed. In that sense, **heat wave** is the most harmful event with 70 casualties.

Since the analysis is focused on the addition of injuries and fatalities, a deeper analysis could find more precision by doing the same process with distinction between casualties. Other calculations could be made to define harmful casualties, like the total amount of casualties by events types or the related proportion.

### Economic consequences

This process analysis is mostly identical to the previous one.

#### Data processing

Subset to get only the working variables :
```{r eco.subset}
econ <- storm[, c("EVTYPE", "PROPDMG", "CROPDMG", "PROPDMGEXP", "CROPDMGEXP")]
```

With related unit prefix variables get the explicit number of economic amount casualties :
```{r eco.prefix, message=FALSE, warning=FALSE}
# Types of prefix
table(econ$PROPDMGEXP)
table(econ$CROPDMGEXP)

# Replace prefix nomination as a related decimal unit multiplication variable
multipre <- function(prefix){
      if (prefix == "" | prefix == 0){
            1
      } else if (prefix == "K" | prefix == "k") {
            10^3
      } else if (prefix == "M" | prefix == "m"){
            10^6
      } else if (prefix == "B" | prefix == "b"){
            10^9
      } else if (!is.na(as.numeric(prefix))){
            as.numeric(prefix)
      } else {
            0
      }
}
econ <- econ %>% mutate(PropMulti = sapply(PROPDMGEXP, multipre), CropMulti = sapply(CROPDMGEXP, multipre))
```

The above code is subjective to interpretation since some prefix in the data set are outside of the range specified by the [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), which is quoted as :  
> Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions.   

Add fatalities and injuries variables count to get total amount of population health consequences :
```{r eco.casualties, message=FALSE}
econ <- econ %>% mutate(Casualties = PROPDMG*PropMulti + CROPDMG*CropMulti)
```

Overview of the casualties variable :
```{r eco.casualties.overview, message=FALSE}
summary(econ$Casualties)

# Proportion of harmless events :
mean(econ$Casualties == 0)

# Without harmless events :
summary(econ$Casualties[econ$Casualties != 0])
```

Now, the data is grouped by types of events and 
```{r eco.casualties.harmful}
# Group by types of events
byEvtype <- econ %>% group_by(EVTYPE)

# Aggregate by types of events with their related maximum number of casualties and proportion of events with at least one casualty
summaryzero <- summarize(byEvtype, harmful = max(Casualties),
                         propHarmEvent = sum(Casualties != 0)/n(),
                         numEvent = n())

# Show the above result by descending order by :
## 1. maximum number of casualties
arrange(summaryzero[summaryzero$harmful !=0,], desc(harmful), desc(propHarmEvent))

## 2.proportion of types of events with at least one casualty
arrange(summaryzero[summaryzero$harmful !=0,],
        desc(propHarmEvent), desc(harmful))
```

#### Conclusion

Same as the previous section, the above results show two different types of answer:  
 
      1. One is dealing with the maximum number of dollars of damage on property and crops, "harmful" variable. In that sense, **flood** events are the most impactful with the biggest event doing 115 032 500 000$ of damage.  
      
      2. The other is dealing with the propotion of types of events with at least one casualty, "popHarmEvent" variable. This result is skewed since most of the bigger proportion variable values are related to few events types occurences. By consequence, "popHarmEvent" top 10 have few type of event observed. In that sense, the aggreagte **tornadoes, tstm wind, hail** is the most harmful event(s) with 1 602 500 000$ of damage.

Since the analysis is focused on the addition of property and crops damages, a deeper analysis could find more precision by doing the same process with distinction between the two variables. Other calculations could be made to define harmful casualties, like the total amount of casualties by events types or the related proportion.