---
title: "Inferential data analysis"
author: "Akim van Eersel"
date: '2020-10-11'
output: 
  pdf_document: 
    fig_height: 4
---

# Exploratory data analysis : tooth growth data

The data set 'ToothGrowth' is aiming to study the effect of vitamin C on tooth growth in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC).

## Load the data

```{r load}
library(datasets)

# Load the data
data(ToothGrowth)

# Store the data set in a variable
tooth <- ToothGrowth

# Rename some values to explicit labeling
tooth$supp <- factor(tooth$supp, labels= c("orange juice", "ascorbic acid"))
colnames(tooth) <- c('len', 'delivery.method', 'dose')
```

## Statistics summary

Get the dimensions of the data set and by features.
```{r dim}
# Number of rows and columns
dim(tooth)

# Number of observations for the "dose" and "delivery.method" variables
table(tooth$dose, tooth$delivery.method)
```

Let's see the 5 first rows of the data set.
```{r head}
head(tooth)
```

Overall statistics on the length of the tooth growth.
```{r fivenum}
# Tukey's five number summary + mean & std
data.frame(Statistic = c('Min', '25th percentile','Median', '75th percentile', 'Max', 'Mean', 'Standard Deviation'),
  Values = c(fivenum(tooth$len), mean(tooth$len), sd(tooth$len)))
```

Now, let's see the length of tooth growth discriminated by other variable in a plot.
```{r plot, message=FALSE}
library(ggplot2)

ggplot(data = tooth, aes(dose, len, color = delivery.method), ) +
      geom_point(size = 3, alpha = 1/2) + 
      geom_smooth(method = "lm", alpha =1/2, show.legend = F)  +
      labs(title = "Tooth growth length by dose and by vitamin C delivery method",
           x = "Dose levels (mg/day)", y = "Tooth growth length")
```

The plot is showing all data points and a linear regression of the length function of the dose levels for each vitamin C delivery method.

# Inferential analysis

## Hypothesis testing

Two hypothesis are set :  
      1. Whatever the delivery method, tooth growth length is increasing as the increase of the dose levels.  
      2. For both 0.5 and 1.0 mg/day dose levels delivery method is more effective with the orange juice.
      
### Length increase with dose levels

In order to reduce the amount of calculations, we'll only focus on dose levels 0.5 and 2.0 mg/day. In that sense :  
      * Null hypothesis, H0 : means of the length variable are equal, mu_2.0 - mu_0.5 = 0
      * Alternative hypothesis, Ha : means of the length variable are different, mu_2.0 - mu_0.5 > 0

we assume the normality of the distribution of length variable on both dose levels, also as the independence between subjects. Similar variation is observed between the two groups.   
A t-test will be conducted.

```{r dose.hyp}

g2 <- tooth$len[tooth$dose == 2] 
g1 <- tooth$len[tooth$dose == 0.5]

t.test(g2-g1, var.equal = TRUE, alternative = "greater")
```

Even if t-test's results explicitly state that the null hypothesis is rejected, the related p-value less than 0.01 and the 95% confidence interval of the means difference not including 0 confirm H0 rejection.  

In conclusion, whatever the delivery method, tooth growth length is higher with a 2.0 mg/day dose level than a 0.5 mg/day dose level.

### Greater length with on delivery method

In order to reduce the amount of calculations, we'll only focus on dose level 1.5 mg/day. Data on the 1.0 mg/day dose level by delivery method overlap on each other but still seems to be enough differentiated, let's investiguate. In that sense :  
      * Null hypothesis, H0 : means of the length variable are equal, mu_orange.juice - mu_ascorbic.acid = 0
      * Alternative hypothesis, Ha : means of the length variable are different, mu_orange.juice - mu_ascorbic.acid > 0

we assume the normality of the distribution of length variable on both dose levels, also as the independence between subjects. Similar variation is observed between the two groups.   
A t-test will be conducted.

```{r delivery.hyp}

g2 <- tooth$len[tooth$dose == 1 & tooth$delivery.method == "orange juice"] 
g1 <- tooth$len[tooth$dose == 1 & tooth$delivery.method == "ascorbic acid"]

t.test(g2-g1, var.equal = TRUE, alternative = "greater")
```

Again, even if t-test's results explicitly state that the null hypothesis is rejected, the related p-value less than 0.01 and the 95% confidence interval of the means difference not including 0 confirm H0 rejection.  

In conclusion, for a 1.0 mg/day dose level, tooth growth length is higher with orange juice vitamin C delivery method than ascorbic acid vitamin C delivery method.